An Improved Method for Coupling Hydrodynamics with Reaction Networks and Nuclear Statistical Equilibrium

### 2021-05-03 simplified_sdc_intro/paper.tex
update the Jacobian

To solve the reaction system implicitly, the ODE solver needs the Jacobian,
$\partial \Rb/\partial \Uc$.  We follow the method of \cite{castro_sdc}
and factor this into two pieces,
\begin{equation}
{\bf J} = \frac{\partial \Rb}{\partial {\bf w }} \frac{\partial {\bf w}}{\partial \Uc}.
\end{equation}

Even though it has no reactive sources, we include $\rho$ in our conservative state
for the purposes of computing the Jacobian (we denote the conserved state used for the
Jacobian $\Uc^\prime$).
Writing this out for two species, $X_\alpha$ and $X_\beta$, we have
\begin{equation}
\Uc^\prime = \left ( \begin{array}{c} \rho \\ \rho X_\alpha \\ \rho X_\beta \\ \rho E \\ \rho e \end{array} \right )
\end{equation}
We take the intermediate state to be ${\bf w} = (\rho, X_\alpha, X_\beta,
K, T)^\intercal$, where $K$ is the kinetic energy:
\begin{equation}
K = \frac{1}{2} u^2
\end{equation}

The Jacobian transformation $\partial \Uc^\prime/\partial {\bf w}$ is:
\begin{equation}
\frac{\partial \Uc}{\partial {\bf w}} = \left (
   \begin{array}{ccccc}
       1 & 0 & 0 & 0 & 0 \\
       X_\alpha & \rho & 0 & 0 & 0 \\
       X_\beta & 0 & \rho & 0 & 0  \\
       \rho e_\rho  + e + K &
                 \rho  e_{X_\alpha} & \rho e_{X_\beta} & \rho &
                 \rho e_T \\
       \rho e_\rho  + e  &
                 \rho  e_{X_\alpha} & \rho e_{X_\beta} & 0 &
                 \rho e_T
     \end{array}\right)
\end{equation}
where we use the following notation for compactness:
\begin{equation}
e_\rho = \dedrd \qquad
e_T = \dedTd \qquad
e_{X_k} = \dedXd
\end{equation}
and write the kinetic energy as $K = u^2/2$.  We get the inverse
(computed via SymPy, see the included Jupyter notebook) is:

\begin{equation}
\renewcommand{\arraystretch}{1.5}
\frac{\partial {\bf w}}{\partial \Uc^\prime} = \left (
  \begin{array}{ccccc}
   1  & 0 & 0 & 0 & 0 \\
   - \frac{X_\alpha}{\rho} & \frac{1}{\rho} & 0 & 0 & 0 \\
   - \frac{X_\beta}{\rho} & 0 & \frac{1}{\rho} & 0 & 0 \\
   - \frac{K}{\rho} & 0 & 0 & \frac{1}{\rho} & -\frac{1}{\rho} \\
   \frac{\sum_k X_{k} e_{X_k} - \rho e_\rho - e}{ \rho e_T} &
    -\frac{e_{X_\alpha}}{\rho e_T} & -\frac{e_{X_\beta}}{\rho e_T} & 0 & \frac{1}{\rho e_T} \\
   \end{array}\right)
\renewcommand{\arraystretch}{1}
\end{equation}

@@

>>>>

To solve the reaction system implicitly, the ODE solver needs the Jacobian,
$\partial \Rb/\partial \Uc$.  We follow the method of \cite{castro_sdc}
and factor this into two pieces,
\begin{equation}
\label{eq:factored_jac}
{\bf J} = \frac{\partial \Rb}{\partial {\bf w }} \frac{\partial {\bf w}}{\partial \Uc}.
\end{equation}

Even though it has no reactive sources, we include $\rho$ in our conservative state
for the purposes of computing the Jacobian (we denote the conserved state used for the
Jacobian $\Uc^\prime$).
Writing this out for two species, $X_\alpha$ and $X_\beta$, we have
\begin{equation}
\Uc^\prime = \left ( \begin{array}{c} \rho \\ \rho X_\alpha \\ \rho X_\beta \\ \rho E \\ \rho e \end{array} \right )
\end{equation}
We take the intermediate state to be ${\bf w} = (\rho, X_\alpha, X_\beta,
K, e)^\intercal$, where $K$ is the kinetic energy:
\begin{equation}
K = \frac{1}{2} u^2
\end{equation}
Note: this intermediate state differs from what we used in \citet{castro_sdc} in that
we are using $e$ instead of $T$ as an independent variable now.  This follows from
the work shown in \citet{strang_rnaas} that explored integrating the Strang
reaction system with an energy equation.  If the network gives derivatives in terms
of temperature, we can convert as:
\begin{equation}
\left . \frac{\partial \phi}{\partial T} \right |_\rho =
  \left . \frac{\partial \phi}{\partial e} \right |_\rho 
  \left . \frac{\partial e}{\partial T} \right |_\rho =
  c_v \left . \frac{\partial \phi}{\partial e} \right |_\rho
\end{equation}
for some quantity $\phi(\rho, T)$.  Here we identified the specific heat
at constant volume, $c_v = \partial e/\partial T|_\rho$.

The Jacobian transformation $\partial \Uc^\prime/\partial {\bf w}$ is:
\begin{equation}
\frac{\partial \Uc}{\partial {\bf w}} = \left (
   \begin{array}{ccccc}
       1 & 0 & 0 & 0 & 0 \\
       X_\alpha & \rho & 0 & 0 & 0 \\
       X_\beta & 0 & \rho & 0 & 0  \\
       e + K &
                 \rho  e_{X_\alpha} & \rho e_{X_\beta} & \rho &
                 \rho \\
       e  &
                 \rho  e_{X_\alpha} & \rho e_{X_\beta} & 0 &
                 \rho
     \end{array}\right)
\end{equation}
where we use the following notation for compactness:
\begin{equation}
e_{X_k} = \dedXd
\end{equation}
We get the inverse
(computed via SymPy, see the included Jupyter notebook) as:
\begin{equation}
\renewcommand{\arraystretch}{1.5}
\frac{\partial {\bf w}}{\partial \Uc^\prime} = \left (
  \begin{array}{ccccc}
   1  & 0 & 0 & 0 & 0 \\
   - \frac{X_\alpha}{\rho} & \frac{1}{\rho} & 0 & 0 & 0 \\
   - \frac{X_\beta}{\rho} & 0 & \frac{1}{\rho} & 0 & 0 \\
   - \frac{K}{\rho} & 0 & 0 & \frac{1}{\rho} & -\frac{1}{\rho} \\
   \frac{\sum_k X_{k} e_{X_k}  - e}{ \rho} &
    -\frac{e_{X_\alpha}}{\rho} & -\frac{e_{X_\beta}}{\rho} & 0 & \frac{1}{\rho} \\
   \end{array}\right)
\renewcommand{\arraystretch}{1}
\end{equation}

@@

While VODE can compute the entire Jacobian, ${\bf J}$ numerically via differencing, we found
that does not give reliable results.  Instead, we compute $\partial {\bf w}/\partial {\bf U}^\prime$ analytically and $\partial {\bf R}/\partial {\bf w}$ numerically via one-sided differencing.
The resulting Jacobian, as computed via equation~\ref{eq:factored_jac}, is larger than
expected by the integrator because of the inclusion of the $\rho$
terms in the intermediate steps, so we remove the row corresponding to
$R(\rho)$ and the column corresponding to $\partial/\partial \rho$
before passing the Jacobian to VODE.
